---
title: "Initial Data Exploration"
output: html_notebook
---

Goal: Having a closer look at the data and doing some sanity checks.

```{r}
library(tidyverse)

pat <- read_delim("Random_PatientLevelInfo_2020.tsv", delim = "\t")
lab <- read_delim("Random_LabValuesInfo_2020.tsv", delim = "\t")
```

## Patients

```{r}
pat
```
 
 The file has one row per subject .

```{r}
length(unique(pat$USUBJID))
length(unique(lab$USUBJID)) # there are lab values for all 452 patients
```
 
 There is only one study in the data set.
 It has three arms with ~150 subjects in each of them.
 
```{r}
pat %>% 
  count(STUDYID)

pat %>% 
  count(ACTARM, ACTARMCD)
```

```{r}
pat %>% 
  count(SEX)
```

I'm assuming "Undifferentiated" is a simple data entry error.

```{r}
pat <- pat %>% 
  mutate(
    SEX = fct_collapse(SEX,
      U = c("UNDIFFERENTIATED", "U")
    )
  )
```

Some of the race categories are almost empty. It might be ok to combine them.

```{r}
pat %>% 
  count(RACE)
```

## Lab values 

```{r}
lab <- lab %>% 
  mutate(
    BMRKR2 = fct_relevel(BMRKR2, c("LOW", "MEDIUM", "HIGH")),
    AVISIT = fct_relevel(AVISIT, "SCREENING", after = 0)
  )
```


ALl measurements are from the same study 

```{r}
lab %>% 
  count(STUDYID)
```

 * ALT is increased with liver damage and is used to screen for and/or monitor liver disease
 * CRP is a blood test marker for inflammation in the body (e.g., chronic inflammatory diseases such as lupus, vasculitis, or rheumatoid arthritis (RA)
 * high IGA levels might be caused by allergies, chronic infections, autoimmune disorders such as RA
 
```{r}
lab
```

There are two biomarkers (no obvious correlation between them). I'm making marker 2 an ordered factor.

```{r}

lab %>% 
  ggplot(aes(BMRKR2, BMRKR1)) +
  geom_boxplot()
```

No surprises here - test names, categories and units match nicely.

```{r}
lab %>% 
  count(LBTEST, LBTESTCD, LBCAT, AVALU)
```

It seems like we have consistent measures, no missing values, no values that are clearly wrong.

```{r fig.asp=0.5}
lab %>% 
  group_by(USUBJID, LBTEST) %>% 
  summarise(n_observations = n()) %>% 
  group_by(LBTEST) %>% 
  count(n_observations)

lab %>% 
  count(AVISIT)

any(is.na(lab$AVAL))

lab %>%  
  ggplot(aes(AVISIT, AVAL)) +
  geom_boxplot() +
  facet_wrap(~LBTEST) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

#### some randomly selcted patient profiles

```{r fig.asp=0.5}

tmp <- pat %>% 
  group_by(ACTARM) %>% 
  sample_n(20) %>% 
  left_join(lab, by = "USUBJID") 

tmp %>%  
  ggplot(aes(AVISIT, AVAL, group = USUBJID)) +
  geom_line(alpha = .1) +
  facet_grid(LBTEST ~ ACTARM) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

### Mean group performance over time, by study arm, Biomarker2

```{r fig.asp=0.5}

tmp <- pat %>% 
  left_join(lab, by = "USUBJID") 

tmp %>%  
  ggplot(aes(AVISIT, AVAL, group = ACTARM, color = ACTARM)) +
  geom_smooth(alpha = .05, method = "loess") +
  facet_wrap(~LBTEST) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

tmp %>%  
  ggplot(aes(AVISIT, AVAL, group = ACTARM, color = ACTARM)) +
  geom_smooth(se = FALSE, method = "loess") +
  facet_grid(BMRKR2 ~ LBTEST) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

### Change relative to baseline, by study arm, Biomarker 2

```{r fig.asp=0.5}

tmp <- pat %>% 
  left_join(lab, by = "USUBJID") %>% 
  filter(AVISIT != "SCREENING") %>% 
  group_by(USUBJID, LBTEST) %>% 
  mutate(AVAL_CHANGE = AVAL - first(AVAL)) %>% 
  ungroup()

tmp %>%  
  ggplot(aes(AVISIT, AVAL_CHANGE, group = USUBJID, color = ACTARM)) +
  geom_line(alpha = .05) +
  geom_smooth(aes(group = 1)) +
  facet_grid(ACTARM~LBTEST) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

tmp %>%  
  ggplot(aes(AVISIT, AVAL_CHANGE, group = ACTARM, color = ACTARM)) +
  geom_smooth(alpha = .05, method = "loess") +
  facet_wrap(~LBTEST) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))

tmp %>%  
  ggplot(aes(AVISIT, AVAL_CHANGE, group = ACTARM, color = ACTARM)) +
  geom_smooth(alpha = .05, method = "loess") +
  facet_wrap(BMRKR2 ~ LBTEST) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```